Global Life Expectancy: Unraveling Health and Economic Determinants (2000-2015)

### 2024-11-30 paper/paper.qmd
initial

2.  **Continent(`Continent`)**: This is the categorical variable showing the geographic information of each country. Continent usually

3.  **Percentage Expenditure on Health (`PercentageExpenditure`)**:

4.  **Total expenditure (`TotalExpenditure`)**:

5.  **Income composition of resource (`IncomeComposition`)**:

6.**Schooling (`Schooling`)**:

@@

#Plot1: create bar plot of count of status
numberrows <- ggplot(cleaned_expectancy, aes(x = Status, fill = Status)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(
    title = "Distribution of Count of Developed and Developing Countries",
    x = "Status (Developed vs Developing)",
    y = "Count"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Developed" = "skyblue", "Developing" = "darkblue")) +
  theme(legend.position = "none",
  plot.title = element_text(size = 7))

# Plot2: Create a bar plot to show the distribution of countries by Status
numbercountry <- ggplot(country_counts, aes(x = Status, y = n, fill = Status)) +
  geom_bar(stat = "identity") +  # Use 'identity' to directly use the count values
  geom_text(aes(label = n), vjust = -0.5) +  # Add count labels on top of bars
  labs(
    title = "Distribution of Countries by Status (Developed vs Developing)",
    x = "Status (Developed vs Developing)",
    y = "Number of Countries"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Developed" = "skyblue", "Developing" = "darkblue")) +
  theme(
    legend.position = "none", 
  plot.title = element_text(size = 7))

@@

```{r}
#| label: fig-continent
#| fig-cap: "Distribution of contient. Africa has the highest number of countries represented, followed by Asia among all 5 continents. "
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

#Plot1: Distribution of number of data per continent
#Create the bar plot showing the number of data per continent
continentrow <- ggplot(cleaned_expectancy, aes(x = Continent)) +
  geom_bar(fill = "orange", color = "white") +
  labs(
    title = "Distribution of Count by Continent",
    x = "Continent",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12)
  ) +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5, size = 2.5)

# Plot2: The distribution of number of countries each continent
#Create the bar plot showing the number of unique countries per continent
continentnumber <-ggplot(continent_count, aes(x = Continent, y = num_countries)) +
  geom_bar(stat = "identity", fill = "#FFA07A", color = "white") +
  labs(
    title = "Number of Countries by Continent",
    x = "Continent",
    y = "Number of Countries"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10)
  ) +
  geom_text(aes(label = num_countries), vjust = -0.5, size = 2.5)

@@

@@

To predict the outcome of the life expectancy of people from different regions, we developed tow linear regression models: one for the developed country and one for developing country. These models aim to estimate the life expectancy of people under different environment.

@@

This paper utilizes linear regression models: $$LifeExpecetancy_I = \beta_0 + \beta_1 $$ $$LifeExpectancy_D = \beta_0 + \beta_1$$

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

>>>>

2.  **Continent(`Continent`)**: This is the categorical variable showing the geographic information of each country. Developed continents, for example Europe and North America, typically have well-established healthcare infrastructure, which are considered to contribute to higher life expectancy. Factors like clean water access, education resources, nutrition and lifestyle factors can significantly impact life expectancy and vary greatly across continents. @fig-continent reveals that Africa has the highest number of countries represented with the largest portion of the data. Other continents, America and Europe, have fewer countries included, with only 9 countries from Oceania.

3.  **Percentage Expenditure on Health (`PercentageExpenditure`)**: This variable reflects the expenditure allocated to health-related expenditures as a percentage of GDP per capita.

4.  **Total expenditure (`TotalExpenditure`)**: This variable shows the general government expenditure on health as a percentage of total government expenditure. This is considered as a reflection of economic situation of a country. A higher percentage indicates that a country places significant emphasis on healthcare as a key area of public investment.

5. **Income composition of resource (`IncomeComposition`)**: his is a numerical variable representing the income component of the Human Development Index (HDI).

6. **Schooling (`Schooling`)**: This is a numerical variable showing the number of education per person.

@@

#Plot1: create bar plot of count of status
numberrows <- ggplot(cleaned_expectancy, aes(x = Status, fill = Status)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5,size = 2.5) +
  labs(
    title = "Distribution of Count of Country by Status",
    x = "Status (Developed vs Developing)",
    y = "Count"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Developed" = "skyblue", "Developing" = "darkblue")) +
  theme(legend.position = "none",
  plot.title = element_text(size = 7,face = "bold"))

# Plot2: Create a bar plot to show the distribution of countries by Status
numbercountry <- ggplot(country_counts, aes(x = Status, y = n, fill = Status)) +
  geom_bar(stat = "identity") +  # Use 'identity' to directly use the count values
  geom_text(aes(label = n), vjust = -0.5,size = 2.5) +  # Add count labels on top of bars
  labs(
    title = "Distribution of Country by Status",
    x = "Status (Developed vs Developing)",
    y = "Number of Countries"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Developed" = "skyblue", "Developing" = "darkblue")) +
  theme(
    legend.position = "none", 
  plot.title = element_text(size = 7,face = "bold"))

@@

```{r}
#| label: fig-continent
#| fig-cap: "Distribution of contient. Africa has the highest number of countries represented, followed by Asia among all 5 continents. The distribution of count of data aslo shows that a significant portion of the data is focused on African and Asian nations.  "
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

#Plot1: Distribution of number of data per continent
#Create the bar plot showing the number of data per continent
continentrow <- ggplot(cleaned_expectancy, aes(x = Continent)) +
  geom_bar(fill = "orange", color = "white") +
  labs(
    title = "Distribution of Count by Continent",
    x = "Continent",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 10)
  ) +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5, size = 2.5)

# Plot2: The distribution of number of countries each continent
#Create the bar plot showing the number of unique countries per continent
continentnumber <-ggplot(continent_count, aes(x = Continent, y = num_countries)) +
  geom_bar(stat = "identity", fill = "#FFA07A", color = "white") +
  labs(
    title = "Distribution of Country by Continent",
    x = "Continent",
    y = "Number of Country"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 10)
  ) +
  geom_text(aes(label = num_countries), vjust = -0.5, size = 2.5)

@@

```{r}
#| label: fig-percentage
#| fig-cap: "Distribution of expenditure allocated to health-related expenditures as a percentage of GDP per capita. Africa has the highest number of countries represented, followed by Asia among all 5 continents. The distribution of count of data aslo shows that a significant portion of the data is focused on African and Asian nations.  "
#| echo: false
#| eval: true
#| warning: false
#| message: false

```{r}
#| label: fig-total
#| fig-cap: "Distribution of contient. Africa has the highest number of countries represented, followed by Asia among all 5 continents. The distribution of count of data aslo shows that a significant portion of the data is focused on African and Asian nations.  "
#| echo: false
#| eval: true
#| warning: false
#| message: false

```{r}
#| label: fig-income
#| fig-cap: "Distribution of contient. Africa has the highest number of countries represented, followed by Asia among all 5 continents. The distribution of count of data aslo shows that a significant portion of the data is focused on African and Asian nations.  "
#| echo: false
#| eval: true
#| warning: false
#| message: false

```{r}
#| label: fig-schooling
#| fig-cap: "Distribution of contient. Africa has the highest number of countries represented, followed by Asia among all 5 continents. The distribution of count of data aslo shows that a significant portion of the data is focused on African and Asian nations.  "
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

To predict the outcome of the life expectancy of people from different regions, we developed two linear regression models using R [@citeR]: one for the developed country and one for developing country. The outcome variable, `LifeExpectancy`, is a continuous and represents the average number of years a person is expected to live, assuming all related resources remain constant throughout their lifetime. These models aim to estimate the life expectancy of people under different environment. The GAM allows for capturing potential non-linear relationships between the predictors and the outcome.

The model is specified as follows:

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```

$$overall\_time\_seconds_i​=\beta_0​+s(age_i​)+\beta_1⋅gender_i​+s(race\_count_i​​)+ \beta_2⋅iaaf\_category​​ + ϵ_i​$$

Where:

-   $overall\_time\_seconds_i$ is the marathon finishing time in seconds for the i-th runner.

-   $s(age_i)$ is a smooth function of the runner’s age at the time of the race.

-   $gender_i$ is the runner's gender (treated as a categorical variable).

-   $s(race\_count_i)$ is a smooth function of the total number of NYRR races the runner has participated in.

-   $iaaf\_category_i$​ is the country the runner is from under IAAF standards.

-   $\epsilon_i$ is the error term, assumed to be normally distributed with a mean of 0 and constant variance $ϵ_i​∼N(0,\sigma^2)$

This GAM model allows for smooth, non-linear effects of continuous predictors (**`age`** and **`races_count`**) while keeping the categorical predictors (**`gender`** and **`iaaf_category`**) in a linear framework.

@@

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
